---
title: "Development underpins the summer solstice reversal of the effects of climate warming on the autumn phenology of European beech"
subtitle: "Analysis of leaf senescence data for Experiment 1"
author: "Dominic Rebindaine"
date: "2025-09-18"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction

This document displays the analysis of leaf chlorophyll data of experiment 1 in "Development underpins the summer solstice reversal of the effects of climate warming on the autumn phenology of European beech" Rebindaine et al. (2025). The data is loaded and restructured, leaf senescence dates (end-of-season, EOS) are derived and analysed in context of the various treatments, then figures are plotted.

### Load required packages

```{r, message = FALSE}
library(tidyverse)
library(cropgrowdays)
library(lmerTest)
require(broom.mixed)
require(patchwork)
require(data.table)
require(lme4)
require(eoffice)
require(ggpmisc)
require(parameters)
library(MuMIn)
library(vegan)
```

### Configure input/output

```{r in/out}

input.path = '/Users/dominic/Library/Mobile Documents/com~apple~CloudDocs/Documents/PhD/development_2022_2023/analysis/experiment_1/data'

output.path = '/Users/dominic/Library/Mobile Documents/com~apple~CloudDocs/Documents/PhD/development_2022_2023/analysis/experiment_1/figures/leaves/new'

```

### Load and restructure data

```{r load data}
# import data
chlor.df = read.csv(paste(input.path, '2023_phen_exp_spad.csv', sep = '/'))

chlor.df = chlor.df %>% 
  # remove second last measurement point as data from 07.12.2023 is anomalous - unreasonably high due to widespread browning, which causes false readings.
  select(-c('SPAD.07.12.2023')) %>% 
  #turn treatment groups into meaningful terms
    mutate(treatment.timing = ifelse(group %in% c('N','P','O','Q'), 'July',
                                   ifelse(group %in% c('R','S','T','U'), 'August', 'Control')),
          cooling.level = ifelse(group %in% c('N','O','R','S'), 'Moderate',
                                ifelse(group %in% c('P','Q','T','U'), 'Extreme', 'Control'))) %>% 
  # Make the df long format
pivot_longer(-c('id', 'group', 'number', 'cooling.level', 'lo.timing', 'treatment.timing', 'lo.doy'),
                           names_to = 'measurement.date', values_to = 'SPAD')

# recode measurement date to date and remove 'SPAD.' string
chlor.df$measurement.date = as.Date(substr(chlor.df$measurement.date,6,15), format = '%d.%m.%Y')

str(chlor.df)

######### factorise and relevel predictors

# leaf-out timing
chlor.df$lo.timing = as.factor(chlor.df$lo.timing)
# treatment timing
chlor.df$treatment.timing = as.factor(chlor.df$treatment.timing)
chlor.df$treatment.timing = fct_relevel(chlor.df$treatment.timing, 'Control', 'July')
# cooling level
chlor.df$cooling.level = as.factor(chlor.df$cooling.level)
chlor.df$cooling.level = fct_relevel(chlor.df$cooling.level, 'Control', 'Moderate')

# remove NAs
chlor.df = chlor.df %>% 
  filter(!is.na(SPAD)) %>% 
 # group to individual 
  group_by(id) %>% 
   # only keep leaves with more than 8 measurements
     filter(n() >= 8) %>% 
      ungroup()

# Add measurement.doy column
chlor.df$measurement.doy = day_of_year(chlor.df$measurement.date)

# remove L217 as it senesced abnormally early
chlor.df = chlor.df %>% filter(id != 'L217')

# Convert SPAD to chlorophyll content
SPADtoChl = function(SPAD){
  Chl = ifelse(SPAD==0, 0, -0.0029*(SPAD^2) + 1.175*SPAD + 3.8506)
  return(Chl)
}
#SPAD: unitless
#Chl: leaf chlorophyll in microgram / gram fresh weight
#Intercept forced to zero

#From Percival et al. 2008: The Potential of a Chlorophyll Content SPAD Meter to Quantify Nutrient Stress in Foliar Tissue of Sycamore (Acer pseudoplatanus), English Oak (Quercus robur), and European Beech (Fagus sylvatica)

chlor.df = chlor.df %>% 
  mutate(chlorophyll = SPADtoChl(SPAD))

# get mean and sd of groups over time
sum_chlor = chlor.df %>% 
  group_by(group, measurement.doy, measurement.date, treatment.timing, lo.timing, cooling.level) %>% 
  summarise(across(chlorophyll, list(
        mean = \(x) mean(x, na.rm = TRUE), 
        sd = \(x) sd(x, na.rm = TRUE))),
    .groups = 'keep')

```


### Outlier detection
Find measurements that are more than 1.5 times the IQR outside of the lowest/highest quartiles. 
Quartiles and IQR are computed within treatment and measurement date.

```{r}

# Function to flag outliers
flag_iqr_outliers = function(x) {
  q1 = quantile(x, 0.25, na.rm = TRUE)
  q3 = quantile(x, 0.75, na.rm = TRUE)
  iqr = q3 - q1
  lower = q1 - 1.5 * iqr
  upper = q3 + 1.5 * iqr
  return(x < lower | x > upper)
}

# Apply the function per treatment and date
chlor_outliers_flagged = chlor.df %>%
  group_by(group, measurement.date) %>%
  mutate(outlier = flag_iqr_outliers(chlorophyll)) %>%
  ungroup()

# visualise outliers

ggplot(chlor_outliers_flagged, aes(x = measurement.date, y = chlorophyll)) +
  geom_point(aes(color = outlier)) +
  geom_text(
    data = subset(chlor_outliers_flagged, outlier == TRUE),
    aes(label = id),
    vjust = -0.5,        # adjust vertical position
    size = 3,            # text size
    color = "black",     # override point color for visibility
    check_overlap = TRUE # avoid overlapping labels
  ) +
  facet_wrap(~ group) +
  labs(
    title = "Outliers by treatment and measurement date (IQR Method)") +
  theme_minimal()

##### Remove trees with > 1 outlier measurement and then delete outlier points

# Count outliers per tree
tree_outlier_counts = chlor_outliers_flagged %>%
  group_by(id) %>%
  summarise(n_outliers = sum(outlier, na.rm = TRUE)) %>%
  ungroup()

# Identify trees with more than 1 outlier
trees_to_remove = tree_outlier_counts %>%
  filter(n_outliers > 1) %>%
  pull(id)

# Remove those trees from the dataset
chlor.df = chlor_outliers_flagged %>%
  filter(!(id %in% trees_to_remove) & !outlier) %>% 
    select(!outlier)

message("Removed trees with >1 outlier: ", paste(trees_to_remove, collapse = ", "))

```


### Sample sizes

```{r sample sizes}
for (date in unique(chlor.df$measurement.date)) {
  
  print(date-19357) # returns date as doy
  # Observed trees per treatment
print(table(chlor.df[chlor.df$measurement.date== date,]$group))
  
}

# Number of individual trees
length(unique(chlor.df$id))
```

### Relative chlorophyll levels

```{r rel chlor}

chlor.df = chlor.df %>%
  group_by(id)%>%
  mutate(
    #get relative chlorophyll content per individual
    chlorophyll.rel = chlorophyll / max(chlorophyll),
    #save max chlorophyll content for each individual
    chlorophyll.max = chlorophyll.rel == max(chlorophyll.rel),
    #get date of maximum chlorophyll content per individual
    date.max.chl = measurement.date[which.max(chlorophyll)]) %>%
  ungroup()

```


### EOS (leaf senescence timing) derivation

```{r EOS}

## Interpolate 10%, 25%, 50% and 85% senescence dates from seasonal chlorophyll measurements

# Interpolate EOS10 (date chlorophyll last dropped below 90%)
senescence.data.10 = chlor.df %>%
  group_by(id, group, lo.doy, cooling.level, lo.timing, treatment.timing) %>%
  # Find the last row with >90% chlorophyll in each group, keep only this row and next
  filter(row_number() <= max(which(chlorophyll.rel > 0.9))+1,
         row_number() >= max(which(chlorophyll.rel > 0.9))) %>%
  #Linear interpolation
  dplyr::summarize(EOS     = as.Date(approx(chlorophyll.rel, measurement.date, .9,  ties=min)$y, origin="1970-01-01")) %>%
  mutate(EOS.DOY    = yday(EOS),
         phenophase = "EOS10") %>%
  ungroup()

# Interpolate EOS25 (date chlorophyll last dropped below 75%)
senescence.data.25 = chlor.df %>%
  group_by(id, group, lo.doy, cooling.level, lo.timing, treatment.timing) %>%
  # Find the last row with >75% chlorophyll in each group, keep only this row and next
  filter(row_number() <= max(which(chlorophyll.rel > 0.75))+1,
         row_number() >= max(which(chlorophyll.rel > 0.75))) %>%
  #Linear interpolation
  dplyr::summarize(EOS     = as.Date(approx(chlorophyll.rel, measurement.date, .75,  ties=min)$y, origin="1970-01-01")) %>%
  mutate(EOS.DOY    = yday(EOS),
         phenophase = "EOS25") %>%
  ungroup()

# Interpolate EOS50 (date chlorophyll last dropped below 50%)
senescence.data.50 = chlor.df %>%
  group_by(id, group, lo.doy, cooling.level, lo.timing, treatment.timing) %>%
  # Find the last row with >50% chlorophyll in each group, keep only this row and next
  filter(row_number() <= max(which(chlorophyll.rel > 0.5))+1,
         row_number() >= max(which(chlorophyll.rel > 0.5))) %>%
  #Linear interpolation
  dplyr::summarize(EOS     = as.Date(approx(chlorophyll.rel, measurement.date, .5,  ties=min)$y, origin="1970-01-01")) %>%
  mutate(EOS.DOY    = yday(EOS),
         phenophase = "EOS50") %>%
  ungroup()

# Interpolate EOS85 (date chlorophyll last dropped below 85%)
senescence.data.85 = chlor.df %>%
  group_by(id, group, lo.doy, cooling.level, lo.timing, treatment.timing) %>%
  # remove measurements on 07.12.23 due to measurement error
  #filter(measurement.date != '2023-12-07') %>% 
  # remove P318 as it never falls below 20% of its peak chlorophyll
  filter(id != 'P318') %>% 
  # Find the last row with >85% chlorophyll in each group, keep only this row and next
  filter(row_number() <= max(which(chlorophyll.rel > 0.15))+1,
         row_number() >= max(which(chlorophyll.rel > 0.15))) %>%
  #Linear interpolation
  dplyr::summarize(EOS     = as.Date(approx(chlorophyll.rel, measurement.date, .15,  ties=min)$y, origin="1970-01-01")) %>%
  mutate(EOS.DOY    = yday(EOS),
         phenophase = "EOS85") %>%
  ungroup()

# Rbind
senescence.data = rbind(senescence.data.10,senescence.data.25,senescence.data.50, senescence.data.85)

# Mean EOS DOY per treatment
##############################

eos.doy.mean = 
  senescence.data %>% 
  group_by(group, phenophase) %>% 
  summarise(mean.EOS.DOY = round(mean(EOS.DOY),0))


# Summary for EOS50
##############################

AverageSummary.EOS50 = senescence.data.50 %>% 
  group_by(group) %>% 
  summarise(
    mean.LO.DOY = round(mean(lo.doy),1),
            ci.lower.LO = round(mean.LO.DOY - 1.96*standard_error(lo.doy),1),
            ci.upper.LO = round(mean.LO.DOY + 1.96*standard_error(lo.doy),1),
    meanEOS.DOY = round(mean(EOS.DOY),1),
            ci.lower.EOS = round(meanEOS.DOY - 1.96*standard_error(EOS.DOY),1),
            ci.upper.EOS = round(meanEOS.DOY + 1.96*standard_error(EOS.DOY),1)
  )

```

### Complete dataframe

```{r}
# Translate lettered groups into treatment descriptions
senescence.data = senescence.data %>% 
  mutate(
    # add column to combine treatment.timing and cooling.level
    treat.cool.combined = paste(treatment.timing, cooling.level,sep = ".")
    )

##factorise and relevel treat.cool.combined
senescence.data$treat.cool.combined = as.factor(senescence.data$treat.cool.combined)
senescence.data$treat.cool.combined = fct_relevel(senescence.data$treat.cool.combined, 'Control.Control', 'July.Moderate', 'July.Extreme', 'August.Moderate', 'August.Extreme')

# save prepared dataframe
write_csv(senescence.data, file = paste(input.path, "senescence_data_experiment1.csv", sep = "/"))

```

### ANOVAs by treatment

```{r}

# vector of phenophases
phenophases = unique(senescence.data$phenophase)

# for loop over all phenophases
for (phenophase in phenophases) {
  
  print(phenophase)
  
  model.anova.phenophase = senescence.data %>% 
    filter(phenophase == phenophase) %>% 
      aov(EOS.DOY ~ group, data=.)
  
  result.anova.phenophase = summary(model.anova.phenophase)
  
  print(result.anova.phenophase)
  
}

```
### ANOVAs by leaf-out timing

```{r}

# for loop over all phenophases
for (phenophase in phenophases) {
  
  print(phenophase)
  
  model.lo.anova.phenophase = senescence.data %>% 
    filter(phenophase == phenophase) %>% 
      aov(EOS.DOY ~ lo.timing, data=.)
  
  result.lo.anova.phenophase = summary(model.lo.anova.phenophase)
  
  print(result.lo.anova.phenophase)
  
}

```

### EOS Analysis

```{r leaf senescence analysis}

### Linear modelling of leaf senescence timing 

## Treatment as only  predictor variable
resultsLM = senescence.data.50 %>% 
  do({model = lm(EOS.DOY ~ group, data=.)  # create the model
  data.frame(tidy(model))}) %>%
  filter(!term %in% c("(Intercept)")) %>% 
  mutate(term = factor(term, levels = c("groupM", "groupN", "groupO", "groupP", "groupQ", "groupR", "groupS", "groupT", "groupU")))

resultsLM = resultsLM %>% mutate(term = c("Late_control_control", "Early_july_Moderate", "Late_july_Moderate", "Early_july_Extreme", "Late_july_Extreme", "Early_august_Moderate", "Late_august_Moderate", "Early_august_Extreme", "Late_august_Extreme"))
       
#Show results
as.data.frame(resultsLM) %>%
  dplyr::select(-c(statistic))%>%
  mutate_if(is.numeric, round, digits=1)

## Leaf-out timing as fixed predictor variable
resultsLM.LO = senescence.data.50 %>% 
  do({model = lm(EOS.DOY ~ lo.timing, data=.)  # create the model
  data.frame(tidy(model))}) %>%
  filter(!term %in% c("(Intercept)")) %>% 
  mutate(SEx2 = std.error*2)

#Show results
as.data.frame(resultsLM.LO) %>%
  dplyr::select(-c(statistic))%>%
  mutate_if(is.numeric, round, digits=2)
```

#### Early groups only model

```{r early only model}
resultsLM.Early = senescence.data.50 %>% 
  filter(lo.timing == 'early') %>% 
do({model = lm(EOS.DOY ~ group, data=.)  # create the model
  data.frame(tidy(model))}) %>%
  filter(!term %in% c("(Intercept)")) %>% 
 mutate(term = factor(term, levels = c("groupN", "groupP", "groupR", "groupT")),
        SEx2 = std.error*2,
        lo.timing = 'early')
  
#Show results
as.data.frame(resultsLM.Early) %>%
  dplyr::select(-c(statistic))%>%
  mutate_if(is.numeric, round, digits=2)

# save results for plotting later
toplot.resultsLM.Early = as.data.frame(resultsLM.Early) %>%
  dplyr::select(-c(statistic))%>%
  mutate_if(is.numeric, round, digits=2)
```

#### Late groups only model

```{r late only model}
resultsLM.Late = senescence.data.50 %>% 
  filter(lo.timing == 'late') %>% 
do({model = lm(EOS.DOY ~ group, data=.)  # create the model
  data.frame(tidy(model))}) %>%
  filter(!term %in% c("(Intercept)")) %>% 
 mutate(term = factor(term, levels = c("groupO", "groupQ", "groupS", "groupU")),
        SEx2 = std.error*2,
        lo.timing = 'late')
  
#Show results
as.data.frame(resultsLM.Late) %>%
  dplyr::select(-c(statistic))%>%
  mutate_if(is.numeric, round, digits=2)   

# save results for plotting later
toplot.resultsLM.Late = as.data.frame(resultsLM.Late) %>%
  dplyr::select(-c(statistic))%>%
  mutate_if(is.numeric, round, digits=2)

# merge early and late dfs for plotting later
toplot.resultsLMEarlyLate = rbind(toplot.resultsLM.Early,toplot.resultsLM.Late)

# relevel factors to put treatment pairs in right order
toplot.resultsLMEarlyLate$term = fct_relevel(toplot.resultsLMEarlyLate$term, 'groupN', 'groupO', 'groupP', 'groupQ', 'groupR', 'groupS', 'groupT', 'groupU')
```

#### L vs M analysis

```{r L vs M}
resultsLM.LvM = senescence.data.50 %>% 
  filter(group %in% c('L','M')) %>% 
do({model = lm(EOS.DOY ~ group, data=.)  # create the model
  data.frame(tidy(model))}) %>%
  filter(!term %in% c("(Intercept)")) %>% 
 mutate(term = factor(term, levels = c("groupM")),SEx2 = std.error*2)
  
       
#Show results
as.data.frame(resultsLM.LvM) %>%
  dplyr::select(-c(statistic))%>%
  mutate_if(is.numeric, round, digits=2)

```

#### N vs O analysis

```{r N vs O}
resultsLM.NvO = senescence.data.50 %>% 
  filter(group %in% c('N','O')) %>% 
do({model = lm(EOS.DOY ~ group, data=.)  # create the model
  data.frame(tidy(model))}) %>%
  filter(!term %in% c("(Intercept)")) %>% 
 mutate(term = factor(term, levels = c("groupO")),SEx2 = std.error*2)
  
       
#Show results
as.data.frame(resultsLM.NvO) %>%
  dplyr::select(-c(statistic))%>%
  mutate_if(is.numeric, round, digits=2)
```

#### P vs Q analysis

```{r P vs Q}
resultsLM.PvQ = senescence.data.50 %>% 
  filter(group %in% c('P','Q')) %>% 
do({model = lm(EOS.DOY ~ group, data=.)  # create the model
  data.frame(tidy(model))}) %>%
  filter(!term %in% c("(Intercept)")) %>% 
 mutate(term = factor(term, levels = c("groupQ")),SEx2 = std.error*2)
  
       
#Show results
as.data.frame(resultsLM.PvQ) %>%
  dplyr::select(-c(statistic))%>%
  mutate_if(is.numeric, round, digits=2)

```

#### R vs S analysis

```{r R vs S}
resultsLM.RvS = senescence.data.50 %>% 
  filter(group %in% c('R','S')) %>% 
do({model = lm(EOS.DOY ~ group, data=.)  # create the model
  data.frame(tidy(model))}) %>%
  filter(!term %in% c("(Intercept)")) %>% 
 mutate(term = factor(term, levels = c("groupS")),SEx2 = std.error*2)
  
       
#Show results
as.data.frame(resultsLM.RvS) %>%
  dplyr::select(-c(statistic))%>%
  mutate_if(is.numeric, round, digits=2)

```

#### T vs U analysis

```{r T vs U}
resultsLM.TvU = senescence.data.50 %>% 
  filter(group %in% c('T','U')) %>% 
do({model = lm(EOS.DOY ~ group, data=.)  # create the model
  data.frame(tidy(model))}) %>%
  filter(!term %in% c("(Intercept)")) %>% 
 mutate(term = factor(term, levels = c("groupU")),SEx2 = std.error*2)
  
       
#Show results
as.data.frame(resultsLM.TvU) %>%
  dplyr::select(-c(statistic))%>%
  mutate_if(is.numeric, round, digits=2)

```

#### Days of autumn delay per day of spring delay

```{r days per day}

##################################################################################
## Linear mixed effects model on Autumn leaf senescence change per day change in leaf-out
##################################################################################

## Leaf-out doy as fixed predictor variable with treatment combination (timing plus cooling level) as a random variable
resultsLM.LOD = senescence.data %>% 
    filter(phenophase == "EOS50") %>% 
     do({model = lme4::lmer(EOS.DOY ~ lo.doy + (1|treat.cool.combined), data=.)  # create the model
       data.frame(tidy(model, effect="fixed"))}) %>%
        filter(!term %in% c("(Intercept)")) %>% 
          mutate(SEx2 = std.error*2)

#Show results
as.data.frame(resultsLM.LOD) %>%
  dplyr::select(-c(statistic))%>%
  mutate_if(is.numeric, round, digits=2)


## Only using control treatments
resultsLM.LOD.Control = senescence.data.50 %>% filter(group %in% c('L','M')) %>% 
  do({model = lm(EOS.DOY ~ lo.doy , data=.)  # create the model
  data.frame(tidy(model))}) %>%
  filter(!term %in% c("(Intercept)")) %>% 
  mutate(SEx2 = std.error*2)

#Show results
as.data.frame(resultsLM.LOD.Control) %>%
  dplyr::select(-c(statistic))%>%
  mutate_if(is.numeric, round, digits=2)


```

## Publication Figures

#### Plot theme

```{r plot theme}

  plotTheme1 = theme(
    legend.position   = "right",
    legend.background = element_rect(fill=NA, linewidth=0.5, linetype="solid"),
    legend.text       = element_text(color="black"),
    panel.background  = element_blank(),
    axis.text         = element_text(colour = "black"),
    panel.border      = element_rect(colour = "black", fill=NA),
    axis.line         = element_line(color = "black"),
    strip.background  = element_rect(fill=NA),
    strip.text        = element_text(colour = 'black'),
    plot.title        = element_text(face="bold",hjust = 0.5))
  
  chlorpal = c('#a6cee3','#1f78b4','#b2df8a','#33a02c','#fb9a99','#e31a1c','#fdbf6f','#ff7f00','#cab2d6','#6a3d9a')
  

```


### Leaf senescence figures

```{r leaf sen pub figs}

## plot only Moderate cooling from combined early and late df using facet_grid spacing with points flipped over y = 0
# collapse groups for facetting
toplot.resultsLMEarlyLate2 = toplot.resultsLMEarlyLate %>% 
  mutate(category = fct_collapse(term, 
                                 July_Moderate = c('groupN','groupO'),
                                 July_Extreme = c('groupP','groupQ'),
                                 August_Moderate = c('groupR','groupS'),
                                 August_Extreme = c('groupT','groupU')))

# plot
LMplotLeafSenModerate = ggplot() + 
  scale_color_manual(values = c('#48b649','#2964af'), name = 'Leaf-out timing')+
  geom_hline(yintercept=0)+
  geom_point(data= filter(toplot.resultsLMEarlyLate2, term %in% c("groupN", "groupO", "groupR", "groupS")), 
             aes(x=term, y=-1*estimate, color=lo.timing), size = 3.5,
             position=position_dodge(3))+
  geom_errorbar(data=filter(toplot.resultsLMEarlyLate2, term %in% c("groupN", "groupO", "groupR", "groupS")), 
                aes(x=term, ymin=-1*estimate-1.96*std.error, ymax=-1*estimate+1.96*std.error, color=lo.timing), 
                width=.2, size = 1.5) +
  coord_cartesian(ylim=c(-10,10))+
  xlab("")+ylab("leaf senescence advancement (days)")+
  
  facet_grid(cols = vars(category), scales = 'free_x', space = 'free_x', switch = 'x')+
  plotTheme1 +
  theme(panel.spacing.x = unit(0,"cm"), axis.text.x = element_blank(), axis.ticks.x = element_blank(), 
        axis.text = element_text(size = 20), axis.title = element_text(size = 22),
        strip.text.x = element_text(size = 22), strip.placement = 'outside',
        legend.position.inside = c(.82, .18), legend.title = element_text(size = 20), legend.text = element_text(size = 20)
       )
LMplotLeafSenModerate

#topptx(figure = LMplotLeafSenModerate, filename = 'LMplotLeafSenModerate.pptx')
 
#Save PDF
#ggsave(filename = "Fig2b_moderate_flipped.pdf", plot = LMplotLeafSenModerate, path = output.path, width = 7, height = 6, useDingbats=FALSE)


####
# plot extreme cooling only from combined early and late df using facet_grid spacing with points flipped over y = 0

LMplotLeafSenExtreme = ggplot() + 
  scale_color_manual(values = c('#48b649','#2964af'), name = 'Leaf-out timing')+
  geom_hline(yintercept=0)+
  geom_point(data= filter(toplot.resultsLMEarlyLate2, term %in% c("groupP", "groupQ", "groupT", "groupU")), 
             aes(x=term, y=-1*estimate, color=lo.timing), size = 3.5,
             position=position_dodge(3))+
  geom_errorbar(data=filter(toplot.resultsLMEarlyLate2, term %in% c("groupP", "groupQ", "groupT", "groupU")), 
                aes(x=term, ymin=-1*estimate-1.96*std.error, ymax=-1*estimate+1.96*std.error, color=lo.timing), 
                width=.2, size = 1.5) +
  coord_cartesian(ylim=c(-10,10))+
  xlab("")+ylab("leaf senescence advancement (days)")+
  
  facet_grid(cols = vars(category), scales = 'free_x', space = 'free_x', switch = 'x')+
  plotTheme1 +
  theme(panel.spacing.x = unit(0,"cm"), axis.text.x = element_blank(), axis.ticks.x = element_blank(), 
        axis.text = element_text(size = 20), axis.title = element_text(size = 22),
        strip.text.x = element_text(size = 22), strip.placement = 'outside',
        legend.position.inside = c(.82, .18), legend.title = element_text(size = 20), legend.text = element_text(size = 20)
       )

LMplotLeafSenExtreme

#topptx(figure = LMplotLeafSenExtreme, filename = 'LMplotLeafSenExtreme.pptx')
#Save PDF
#ggsave(filename = "SuppFig_extreme_flipped.pdf", plot = LMplotLeafSenExtreme, path = output.path, width = 7, height = 6, useDingbats=FALSE)

 ################################################################

################################
# plot from all treatments from combined early and late df using facet_grid spacing
# collapse groups for facetting

# plot
LMplotLeafSen2 = ggplot() + 
  scale_color_manual(values = c('#48b649','#2964af'))+
  geom_hline(yintercept=0)+
  geom_point(data= filter(toplot.resultsLMEarlyLate2, term %in% c("groupN", "groupO", "groupP", "groupQ", "groupR", "groupS", "groupT", "groupU")), 
             aes(x=term, y=estimate*-1, color=lo.timing), size = 3.5,
             position=position_dodge(3))+
  geom_errorbar(data=filter(toplot.resultsLMEarlyLate2, term %in% c("groupN", "groupO", "groupP", "groupQ", "groupR", "groupS", "groupT", "groupU")), 
                aes(x=term, ymin=-1*estimate-1.96*std.error, ymax=-1*estimate+1.96*std.error, color=lo.timing), size = 1.5, 
                width=.2, position=position_dodge(.9)) +
  coord_cartesian(ylim=c(-10,10))+
  xlab("")+ylab("Leaf senescence advancement (days)")+
  labs(color = "Leaf-out \n timing") +
  facet_grid(cols = vars(category), scales = 'free_x', space = 'free_x', switch = "x")+
  plotTheme1 +
  theme(panel.spacing.x = unit(-1,"cm"), axis.text = element_text(size = 20), axis.title = element_text(size = 22), axis.text.x = element_blank(), axis.ticks.x = element_blank(), panel.border = element_blank(), legend.position = "inside", legend.position.inside = c(.25, .80), legend.title = element_text(size = 20), legend.text = element_text(size = 20), strip.text.x = element_text(size = 22))

LMplotLeafSen2
#topptx(figure = LMplotLeafSen2, filename = 'LMplotLeafSen2.pptx')
################################

## Early and late treatment pairs e.g. early_control_control vs late_control_control
# LvsM
LMplot90.LvM = ggplot() + 
  scale_color_manual(values = 'dodgerblue3')+
  geom_hline(yintercept=0)+
  geom_point(data= filter(resultsLM.LvM, term %in% c('groupM')), 
             aes(x=term, y=estimate*-1, color=term),size = 3.5,)+
  geom_errorbar(data=filter(resultsLM.LvM, term %in% c('groupM')), 
                aes(x=term, ymin=-1*estimate-1.96*std.error, ymax=-1*estimate+1.96*std.error, color=term), 
                width=.2, size = 1.5) +
  coord_cartesian(ylim=c(-10,10))+
   scale_x_discrete(labels = 'Control') +
  xlab("")+ylab("leaf senescence advancement (days)")+
  plotTheme1 +
   theme(legend.position = "none", panel.border = element_blank(), axis.text = element_text(size = 18), axis.title.y = element_text(size = 20))

LMplot90.LvM

# NvsO
LMplot90.NvO = ggplot() + 
 scale_color_manual(values = 'dodgerblue3')+
  geom_hline(yintercept=0)+
  geom_point(data= filter(resultsLM.NvO, term %in% c('groupO')), 
             aes(x=term, y=estimate*-1, color=term),size = 3.5)+
  geom_errorbar(data=filter(resultsLM.NvO, term %in% c('groupO')), 
                aes(x=term, ymin=-1*estimate-1.96*std.error, ymax=-1*estimate+1.96*std.error, color=term), 
                width=.2, size = 1.5) +
  coord_cartesian(ylim=c(-10,10))+
  scale_x_discrete(labels = 'July_Moderate') +
  xlab("")+ylab("")+
  plotTheme1 +
   theme(legend.position = "none", axis.line.y = element_blank(), axis.text.y = element_blank(), panel.border = element_blank(), axis.ticks.y = element_blank(), axis.text = element_text(size = 18))

LMplot90.NvO

# PvsQ
LMplot90.PvQ = ggplot() + 
 scale_color_manual(values = 'dodgerblue3')+
  geom_hline(yintercept=0)+
  geom_point(data= filter(resultsLM.PvQ, term %in% c('groupQ')), 
             aes(x=term, y=estimate*-1, color=term),size = 3.5)+
  geom_errorbar(data=filter(resultsLM.PvQ, term %in% c('groupQ')), 
                aes(x=term, ymin=-1*estimate-1.96*std.error, ymax=-1*estimate+1.96*std.error, color=term), 
                width=.2,size = 1.5) +
  coord_cartesian(ylim=c(-10,10))+
  scale_x_discrete(labels = 'July_Extreme') +
  xlab("")+ylab("")+
  plotTheme1 +
   theme(legend.position = "none") +theme(legend.position = "none", axis.line.y = element_blank(), axis.text.y = element_blank(), panel.border = element_blank(), axis.ticks.y = element_blank(), axis.text = element_text(size = 18))

LMplot90.PvQ

# RvsS
LMplot90.RvS = ggplot() + 
 scale_color_manual(values = 'dodgerblue3')+
  geom_hline(yintercept=0)+
  geom_point(data= filter(resultsLM.RvS, term %in% c('groupS')), 
             aes(x=term, y=estimate*-1, color=term),size = 3.5)+
  geom_errorbar(data=filter(resultsLM.RvS, term %in% c('groupS')), 
                aes(x=term, ymin=-1*estimate-1.96*std.error, ymax=-1*estimate+1.96*std.error, color=term), 
                width=.2,size = 1.5) +
  coord_cartesian(ylim=c(-10,10))+
  scale_x_discrete(labels = 'August_Moderate') +
  xlab("")+ylab("")+
  plotTheme1 +
   theme(legend.position = "none") +
  theme(legend.position = "none", axis.line.y = element_blank(), axis.text.y = element_blank(), panel.border = element_blank(), axis.ticks.y = element_blank(), axis.text = element_text(size = 18))

LMplot90.RvS

# TvsU
LMplot90.TvU = ggplot() + 
 scale_color_manual(values = 'dodgerblue3')+
  geom_hline(yintercept=0)+
  geom_point(data= filter(resultsLM.TvU, term %in% c('groupU')), 
             aes(x=term, y=estimate*-1, color=term),size = 3.5)+
  geom_errorbar(data=filter(resultsLM.TvU, term %in% c('groupU')), 
                aes(x=term, ymin=-1*estimate-1.96*std.error, ymax=-1*estimate+1.96*std.error, color=term), 
                width=.2,size = 1.5) +
  coord_cartesian(ylim=c(-10,10))+
    scale_x_discrete(labels = 'August_Extreme') +
  xlab("")+ylab("")+
  plotTheme1 +
   theme(legend.position = "none") +
  theme(legend.position = "none", axis.line.y = element_blank(), axis.text.y = element_blank(), panel.border = element_blank(), axis.ticks.y = element_blank(), axis.text = element_text(size = 18))

LMplot90.TvU
################################

# leaf senescence by leaf-out timing boxplot
 lo.eos.boxplot = 
senescence.data.50 %>% ggplot(aes(lo.timing, EOS, fill = lo.timing)) + 
  geom_boxplot(notch = T) + 
  scale_fill_manual(values = c('#48b649','#2964af')) +
   xlab('')+ylab("leaf senescence date") +
  plotTheme1 + theme(legend.position = "none", axis.text = element_text(size = 20), axis.title = element_text(size = 22), axis.text.y = element_text(angle = 15))

 lo.eos.boxplot

#######################################
## Save plots
#######################################

# Early vs Late LeafSen boxplot
#topptx(figure = lo.eos.boxplot, filename = 'early_vs_late_boxplot.pptx')
 
 #ggsave(filename = 'LeafSenBoxplot.pdf', plot = lo.eos.boxplot, path = output.path, height = 6, width = 7)

# Merge separated leaf senescence anomaly plots with each treatment compared to its respective control as well as early vs late boxplot
 
 layout = "AAA
 BBB"
 
 CombinedBox_AnomalyPlot = lo.eos.boxplot + LMplotLeafSenModerate +
   plot_layout(design = layout, tag_level = 'new') + plot_annotation(tag_levels = c('a','b')) &
  theme(plot.tag = element_text(face = 'bold'))
 
  CombinedBox_AnomalyPlot

#####
  # save anomaly plot of all early and late treatments
  ggsave(plot = LMplotLeafSen2, path = output.path, 
         filename = "SuppFig_LeafSenescenceAnomalies.pdf", h = 7, w = 11)
  
  
######################################################################################

#Merge anomaly plots
#define plot layout
layout2 <- "ABCDE"

CombinedchlorSepPlotEarlyLate2 = LMplot90.LvM + LMplot90.NvO + LMplot90.PvQ + LMplot90.RvS + LMplot90.TvU +
  plot_layout(design = layout2, tag_level = 'new')
  
CombinedchlorSepPlotEarlyLate2
#topptx(figure= CombinedchlorSepPlotEarlyLate2, filename = 'CombinedchlorSepPlotEarlyLate2.pptx')

#ggsave(filename = 'SuppFig_all_treatments.pdf', plot = CombinedchlorSepPlotEarlyLate2, path = output.path, height = 6, width = 11)

```

### Relative chlor figures

#### By group

```{r}
# chlor curves
###################

#move points .05 to the left and right
pd = position_dodge(2) 


# Relative chlor 
Relchlor_curve = chlor.df %>%
  mutate(Treatment = factor(group, 
  levels=c("L", "M", "N", "O", "P", "Q", "R", "S", "T", "U"))) %>% 
  ggplot(aes(x=measurement.date, y=chlorophyll.rel*100, colour=Treatment, group=Treatment)) +
  
  geom_hline(yintercept = 15, colour="lightgrey")+
  geom_hline(yintercept = 50, colour="lightgrey")+
  geom_hline(yintercept = 75, colour="lightgrey")+
  geom_hline(yintercept = 90, colour="lightgrey")+
  geom_hline(yintercept = 100, colour="lightgrey")+
  
  geom_vline(xintercept = as.Date('2023-11-01'), colour="lightgrey")+
  geom_vline(xintercept = as.Date('2023-10-01'), colour="lightgrey")+
  geom_vline(xintercept = as.Date('2023-09-01'), colour="lightgrey")+
  geom_vline(xintercept = as.Date('2023-08-01'), colour="lightgrey")+
  
  stat_summary(fun.data = "mean_se", geom="line", size = 1.2, position=pd, alpha=0.7) +
  stat_summary(fun.data = "mean_se", geom="errorbar", size = 0.8, width=0, position=pd) +
  stat_summary(fun.data = "mean_se", geom="point", size = 1.2, position=pd) +
  
  scale_color_manual(values = chlorpal)+
  
  coord_cartesian(ylim=c(0,100),xlim=c(as.Date(c('2023-07-04','2023-12-16'))))+
  
  xlab("") +
  ylab("Relative chlorophyll content (%)") +
  
  #ggtitle("Relative chlor ")+
  
  plotTheme1

Relchlor_curve

```

#### By individual

```{r}

# plot relative chlor length for all individs
individ.relchlor.curves = chlor.df %>% ggplot(
  aes(x=measurement.date,y=chlorophyll.rel*100,colour = group, fill = group)) +
  
  geom_vline(xintercept = as.Date('2023-11-01'), colour="lightgrey")+
  geom_vline(xintercept = as.Date('2023-10-01'), colour="lightgrey")+
  geom_vline(xintercept = as.Date('2023-09-01'), colour="lightgrey")+
  geom_vline(xintercept = as.Date('2023-08-01'), colour="lightgrey")+
  geom_vline(xintercept = as.Date('2023-07-01'), colour="lightgrey")+
  
  geom_hline(yintercept = 15, colour="lightgrey")+
  geom_hline(yintercept = 50, colour="lightgrey")+
  geom_hline(yintercept = 75, colour="lightgrey")+
  geom_hline(yintercept = 90, colour="lightgrey")+
  geom_hline(yintercept = 100, colour="lightgrey")+

  
  geom_line(aes(x=measurement.date, y=chlorophyll.rel*100), color="black")+
  geom_point()+
  
  geom_vline(data=senescence.data.50, 
             aes(xintercept = EOS), linetype = "dashed", colour = 'blue')+
  
  coord_cartesian(ylim=c(0,110),xlim=c(as.Date(c('2023-07-02','2023-12-16'))))+
  
  xlab("Date") +
  ylab("Relative chlor (%)") +
    facet_wrap(~id, dir='v', nrow = 15) +
    plotTheme1 +
  theme(legend.position = "none")
  
individ.relchlor.curves
  #Save PDF
#ggsave(paste(output.path, "individv_chlor_curves.pdf", sep = "/"), height = 40, width = 40)
  

```

## Reproducibility

```{r}

sessionInfo()

```
