---
title: "Experiment_2_bud_analysis_V2"
author: "Dominic Rebindaine"
date: "2025-09-22"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction

This document displays the analysis of bud length data of experiment 2 in "Development underpins the summer solstice reversal of the effects of climate warming on the autumn phenology of European beech" Rebindaine et al. (2025). The data is loaded and restructured, bud set dates (end-of-season, EOS) are derived and analysed in the context of the various treatments, then figures are plotted.

### Load required packages

```{r, message=FALSE}
require(ggplot2)
require(tidyverse)
require(broom.mixed)
require(patchwork)
require(lubridate)
require(data.table)
require(tm)
require(lme4)
require(parameters)
require(ggpmisc)
```

### Configure input/output

```{r in/out}
data.dir = '/Users/dominic/Library/Mobile Documents/com~apple~CloudDocs/Documents/PhD/development_2022_2023/analysis/experiment_2/Data'

output.dir = '/Users/dominic/Library/Mobile Documents/com~apple~CloudDocs/Documents/PhD/development_2022_2023/analysis/experiment_2/figures'

```

### Load and restructure data

```{r load data}

bud.df = read.table(paste(data.dir,"bud_measurements_rm_2022.csv",sep="/"), header=T, sep=",") %>% 
  
  #create ID and treatment columns
  mutate(ID=readr::parse_number(TreeID), #keep only numbers in string
         TreatmentLetter=removeNumbers(TreeID),#remove numbers in string
         date = as.Date(date, format="%d.%m.%Y"),
         #merge controls
         Treatment = ifelse(TreatmentLetter %in% c("A","C", "D","H"),"Control",
                            ifelse(TreatmentLetter %in% c("E"),"Pre_day", 
                                ifelse(TreatmentLetter %in% c("F"),"Pre_night",
                                   ifelse(TreatmentLetter %in% c("G"),"Pre_full", 
                                          ifelse(TreatmentLetter %in% c("I"),"Post_day",
                                                 ifelse(TreatmentLetter %in% c("J"),"Post_night",
                                   ifelse(TreatmentLetter %in% c("K"), "Post_full", "Other")))))))) 
  
  #long format
  bud.df = pivot_longer(bud.df, -c(TreeID, ID, Treatment, TreatmentLetter, week, doy, date), 
               names_to = "bud_type", values_to = "bud_length") %>%
    # remove .length from bud type
  mutate(bud_type = gsub(".length","", bud_type)) %>% 
    # remove B treatment, and NAs
  filter(!Treatment %in% c('Other'),
         !is.na(bud_length)) %>% 
    # group to individ bud
  group_by(ID, bud_type) %>% 
   # only keep buds with more than 8 measurements
  filter(n() >= 8) %>% 
  ungroup()

```


### Sample sizes

```{r sample sizes}

# Observed buds per treatment
table(bud.df[bud.df$date=="2022-09-01",]$Treatment, bud.df[bud.df$date=="2022-09-01",]$bud_type)

# Number of individual trees
length(unique(bud.df$ID))

# Number of individual trees per treatment
length(unique(bud.df[bud.df$Treatment=="Control",]$ID))

length(unique(bud.df[bud.df$Treatment=="Pre_day",]$ID))

length(unique(bud.df[bud.df$Treatment=="Pre_night",]$ID))

length(unique(bud.df[bud.df$Treatment=="Pre_full",]$ID))

length(unique(bud.df[bud.df$Treatment=="Post_day",]$ID))

length(unique(bud.df[bud.df$Treatment=="Post_night",]$ID))

length(unique(bud.df[bud.df$Treatment=="Post_full",]$ID))


```



### Compute bud growth and relative bud lengths

```{r bud growth}

####################################################
## Get absolute and relative bud growth
####################################################


bud.df = bud.df %>%
  group_by(ID, bud_type)%>%
  mutate(
    #get relative bud length per individual
    bud.rel = bud_length / max(bud_length),
    #get bud growth rates per individual
    across(bud_length, ~c(NA, diff(.)), .names = "bud_growth"),
    #get relative bud growth rates per individual
    across(bud.rel, ~c(NA, diff(.)), .names = "relbud_growth"),
    #get date of maximum bud length per individual
    date.max.bud = date[which.max(bud_length)]) %>%
  ungroup()


######################################################################
## Mean abs and rel bud growth per treatment
######################################################################

minbud.df = bud.df %>% 
  group_by(ID, bud_type) %>% 
  filter(row_number() <= which.min(bud_length)) %>% 
  select(ID, Treatment, bud_type, bud_length, bud.rel, date.max.bud)

maxbud.df = bud.df %>% 
  group_by(ID, bud_type) %>% 
  filter(row_number() == min(which(bud.rel > 0.9))) %>% 
  select(ID, Treatment, bud_type, bud_length, bud.rel, date.max.bud)

minmaxbud.df = merge(minbud.df, maxbud.df, by = c('ID', 'bud_type'))

minmaxbud.df = minmaxbud.df %>% 
  group_by(ID, bud_type) %>% 
  mutate(diff = bud_length.y - bud_length.x,
         total.rel.growth = (1-bud_length.x/bud_length.y)*100)

######################################################################
## Get buds that are >= 90% their total length at first measurement
######################################################################

over90start = bud.df %>% filter(doy == 237) %>% filter(bud.rel >= 0.9)

```


### Derive bud set dates and summarise

``` {r}
######################################################################
## Interpolate 90% bud set dates from seasonal bud length measurements
######################################################################

bud.data.90 = bud.df %>%
  group_by(ID, Treatment, bud_type) %>%
  # remove buds that were >= 90% relative bud length on day 1
  filter(!(ID %in% c(116) & bud_type == 'apical'),
         !(ID %in% c(148) & bud_type == 'apical'),
         !(ID %in% c(149) & bud_type == 'apical'),
         !(ID %in% c(156) & bud_type == 'apical'),
         !(ID %in% c(158) & bud_type == 'apical'),
         !(ID %in% c(58) & bud_type == 'lateral'),
         !(ID %in% c(80) & bud_type == 'lateral'),
         !(ID %in% c(82) & bud_type == 'lateral'),
         !(ID %in% c(99) & bud_type == 'lateral'),
         !(ID %in% c(147) & bud_type == 'lateral'),
         !(ID %in% c(153) & bud_type == 'lateral'),
         !(ID %in% c(154) & bud_type == 'lateral'),
         !(ID %in% c(156) & bud_type == 'lateral'),
         !(ID %in% c(157) & bud_type == 'lateral'),
         !(ID %in% c(158) & bud_type == 'lateral'),
         !(ID %in% c(159) & bud_type == 'lateral')) %>% 
  # Find the last row with <90% bud length in each group, keep only this row and next
   filter(row_number() <= min(which(bud.rel > 0.9)),
          row_number() >= min(which(bud.rel > 0.9))-1) %>%
  #Linear interpolation
  summarize(EOS     = as.Date(approx(bud.rel, date, .9,  ties=min)$y, origin="1970-01-01")) %>%
  mutate(EOS.DOY    = yday(EOS)) %>%
  ungroup()

# save data 
write_csv(bud.data.90, file = paste(data.dir, "bud_set_data_experiment2.csv", sep = "/"))

# Observed buds per treatment after removal
table(bud.data.90$Treatment, bud.data.90$bud_type)


# Buds analysis data summary
##############################
BudsSummary = merge(bud.data.90, minmaxbud.df, by = c('ID', 'bud_type')) %>% 
  select(ID, bud_type, Treatment, EOS, EOS.DOY, bud_length.x, bud_length.y, diff, total.rel.growth)

AverageSummary = BudsSummary %>% 
  group_by(Treatment) %>% 
  summarise(
    meanEOS.DOY = round(mean(EOS.DOY),1),
            ci.lower.EOS = round(meanEOS.DOY - 1.96*standard_error(EOS.DOY),1),
            ci.upper.EOS = round(meanEOS.DOY + 1.96*standard_error(EOS.DOY),1),
    meandiff = round(mean(diff),2),
            ci.lower.diff = round(meandiff - 1.96*standard_error(diff),2),
            ci.upper.diff = round(meandiff + 1.96*standard_error(diff),2),
    mean.total.rel = round(mean(total.rel.growth),2),
            ci.lower.rel = round(mean.total.rel - 1.96*standard_error(total.rel.growth),2),
            ci.upper.rel = round(mean.total.rel + 1.96*standard_error(total.rel.growth),2))

AverageSummary

```


## Analysis

### EOS analysis

```{r Bud set analysis}

# Bud set dates
###############

resultsLM = bud.data.90 %>% 
  do({model = lm(EOS.DOY ~ Treatment + bud_type, data=.)  # create the model
  data.frame(tidy(model))}) %>%
  filter(!term %in% c("(Intercept)")) %>% 
  mutate(term = factor(term, levels = c("TreatmentPre_day", "TreatmentPre_night", "TreatmentPre_full",
                                        "TreatmentPost_day", "TreatmentPost_night", "TreatmentPost_full", "bud_typelateral")),
         SEx2 = std.error*2)
       
#Show results
as.data.frame(resultsLM) %>%
  dplyr::select(-c(statistic))%>%
  mutate_if(is.numeric, round, digits=2)

# drop bud type term for plotting later
resultsLM = resultsLM %>% 
  filter(term != "bud.typelateral")

```


### Bud growth analysis

``` {r}
# Autumn bud growth rates
#########################

resultsRelGrowth = bud.df %>% 
  do({model = lmer(relbud_growth*100/7 ~ Treatment + bud_type + (1|date), data=.)  # create your model
  data.frame(tidy(model, effect="fixed"))}) %>% 
  filter(!term %in% c("(Intercept)")) %>% 
  mutate(term = factor(term, levels = c("TreatmentPre_day", "TreatmentPre_night", "TreatmentPre_full",
                                        "TreatmentPost_day", "TreatmentPost_night", "TreatmentPost_full", "bud_typelateral")))
      
#Show results
as.data.frame(resultsRelGrowth) %>%
  dplyr::select(-c(statistic))%>%
  mutate_if(is.numeric, round, digits=2)

# drop bud type term for plotting later
resultsRelGrowth = resultsRelGrowth %>% 
  filter(term != "bud_typelateral")

# Relative Autumn bud growth
#############################

BudGrowth.df = bud.df %>% 
  group_by(Treatment, ID, bud_type) %>% 
  summarize(RelAutGrowth = (1-(min(bud_length)/max(bud_length)))*100) %>% 
  ungroup()

resultsRelAutGrowth = BudGrowth.df %>% 
  do({model = lm(RelAutGrowth ~ Treatment + bud_type, data=.)
  data.frame(tidy(model))}) %>% 
filter(!term %in% c("(Intercept)")) %>% 
  mutate(term = factor(term, levels = c("TreatmentPre_day", "TreatmentPre_night", "TreatmentPre_full",
                                        "TreatmentPost_day", "TreatmentPost_night", "TreatmentPost_full", "bud_typelateral")),
         SEx2 = std.error*2)

#Show results
Intercept = resultsRelAutGrowth[1,]$estimate
as.data.frame(resultsRelAutGrowth) %>%
  #get relative autumn growth change
  mutate(PercentGrowthChange = estimate/Intercept*100,
         PercentGrowth.se = std.error/Intercept*100) %>% 
  dplyr::select(-c(statistic))%>%
  mutate_if(is.numeric, round, digits=2)

# drop bud type term for plotting later
resultsRelAutGrowth = resultsRelAutGrowth %>% 
  filter(term != "bud_typelateral")

# Absolute Autumn bud growth
#############################

resultsAbsAutGrowth = BudsSummary %>% 
  do({model = lm(diff ~ Treatment + bud_type, data=.)
  data.frame(tidy(model))}) %>% 
  filter(!term %in% c("(Intercept)")) %>% 
  mutate(term = factor(term, levels = c("TreatmentPre_day", "TreatmentPre_night", "TreatmentPre_full",
                                        "TreatmentPost_day", "TreatmentPost_night", "TreatmentPost_full", "bud_typelateral")),
         SEx2 = std.error*2)

Intercept = resultsAbsAutGrowth[1,]$estimate
as.data.frame(resultsAbsAutGrowth) %>%
  #get absolute autumn growth change
  mutate(PercentGrowthChange = estimate/Intercept*100,
         PercentGrowth.se = std.error/Intercept*100) %>% 
  dplyr::select(-c(statistic))%>%
  mutate_if(is.numeric, round, digits=2)

# drop bud type term for plotting later
resultsAbsAutGrowth = resultsAbsAutGrowth %>% 
  filter(term != "bud_typelateral")

```


## Figures

### Plot theme

```{r}

  plotTheme1 = theme(
    legend.position   = "none",
    legend.background = element_rect(fill=NA, linewidth=0.5, linetype="solid"),
    legend.text       = element_text(color="black"),
    panel.background  = element_blank(),
    axis.text         = element_text(colour = "black"),
    panel.border      = element_rect(colour = "black", fill=NA),
    axis.line         = element_line(color = "black"),
    strip.background  = element_rect(fill=NA),
    strip.text        = element_text(colour = 'black'),
    plot.title        = element_text(face="bold",hjust = 0.5))


```


### EOS figure

```{r}
# Bud set dates
###############

# Night-Full-Day order mirrored across y = 0
################################################

# relevel resultsLM terms
resultsLM$term = fct_relevel(resultsLM$term, 'TreatmentPre_night', 'TreatmentPre_full', 'TreatmentPre_day', 
                             'TreatmentPost_night','TreatmentPost_full', 'TreatmentPost_day')

## Pre-solstice cooling

LMplot90.pre2 = ggplot() + 
  scale_color_manual(values = c('#F21A00', '#F21A00','#F21A00'))+
  geom_hline(yintercept=0)+
  geom_point(data= filter(resultsLM, term %in% c('TreatmentPre_night', 'TreatmentPre_full', 'TreatmentPre_day')), 
             aes(x=term, y=estimate*-1, color=term), size = 3.5)+
  geom_errorbar(data=filter(resultsLM, term %in% c('TreatmentPre_night', 'TreatmentPre_full', 'TreatmentPre_day')), 
                aes(x=term, ymin=-1*estimate-1.96*std.error, ymax=-1*estimate+1.96*std.error, color=term), linewidth = 1, 
                width=.2, position=position_dodge(.9)) +
  coord_cartesian(ylim=c(-10,10))+
  xlab("")+ylab("Bud set advancement (days)")+
  scale_x_discrete(labels = c('Night','Full', 'Day'))+
  plotTheme1 +
  ggtitle('Pre-solstice cooling')

LMplot90.pre2

## Post-solstice cooling

LMplot90.post2 = ggplot() + 
  scale_color_manual(values = c('#3B9AB2', '#3B9AB2','#3B9AB2'))+
  geom_hline(yintercept=0)+
  geom_point(data= filter(resultsLM, term %in% c('TreatmentPost_night', 'TreatmentPost_full', 'TreatmentPost_day')), 
             aes(x=term, y=estimate*-1, color=term), size = 3.5)+
  geom_errorbar(data=filter(resultsLM, term %in% c('TreatmentPost_night', 'TreatmentPost_full', 'TreatmentPost_day')), 
                aes(x=term, ymin=-1*estimate-1.96*std.error, ymax=-1*estimate+1.96*std.error, color=term), linewidth = 1,
                width=.2, position=position_dodge(.9)) +
  coord_cartesian(ylim=c(-10,10))+
  xlab("")+ylab("Bud set advancement (days)")+
  scale_x_discrete(labels = c('Night','Full', 'Day'))+
  plotTheme1 +
  ggtitle('Post-solstice cooling') +
  theme(axis.title.y=element_blank(),
        axis.text.y=element_blank())

LMplot90.post2


```


### Relative bud growth anomaly

```{r}
# Relative bud growth anomaly
#############################

# relevel resultsRelAutGrowth terms
resultsRelAutGrowth$term = fct_relevel(resultsRelAutGrowth$term, 'TreatmentPre_night', 'TreatmentPre_full', 'TreatmentPre_day', 
                             'TreatmentPost_night','TreatmentPost_full', 'TreatmentPost_day')

## Pre-solstice cooling

LMplotRelGrowthPre = ggplot() + 
  scale_color_manual(values = c('#F21A00', '#F21A00','#F21A00'))+
  geom_hline(yintercept=0)+
  geom_point(data= filter(resultsRelAutGrowth, term %in% c('TreatmentPre_night', 'TreatmentPre_full', 'TreatmentPre_day')), 
             aes(x=term, y=estimate, color=term), size = 3.5)+
  geom_errorbar(data=filter(resultsRelAutGrowth, term %in% c('TreatmentPre_night', 'TreatmentPre_full', 'TreatmentPre_day')), 
                aes(x=term, ymin=estimate-1.96*std.error, ymax=estimate+1.96*std.error, color=term), 
                width=.2, linewidth = 1.5) +
  coord_cartesian(ylim=c(-15,10))+
  xlab("")+ylab("Relative bud growth anomaly (%)")+
  scale_x_discrete(labels = c('Night','Full-day', 'Day'))+
  plotTheme1 +
  ggtitle('Pre-solstice cooling')

LMplotRelGrowthPre

## Post-solstice cooling

LMplotRelGrowthPost = ggplot() + 
  scale_color_manual(values = c('#3B9AB2', '#3B9AB2','#3B9AB2'))+
  geom_hline(yintercept=0)+
  geom_point(data= filter(resultsRelAutGrowth, term %in% c('TreatmentPost_night', 'TreatmentPost_full', 'TreatmentPost_day')), 
             aes(x=term, y=estimate, color=term), size = 3.5)+
  geom_errorbar(data=filter(resultsRelAutGrowth, term %in% c('TreatmentPost_night', 'TreatmentPost_full', 'TreatmentPost_day')), 
                aes(x=term, ymin=estimate-1.96*std.error, ymax=estimate+1.96*std.error, color=term),
                width=.2, linewidth = 1.5) +
  coord_cartesian(ylim=c(-15,10))+
  xlab("")+ylab("Relative bud growth anomaly (%)")+
  scale_x_discrete(labels = c('Night','Full-day', 'Day'))+
  plotTheme1 +
  ggtitle('Post-solstice cooling') +
  theme(axis.title.y=element_blank(),
        axis.text.y=element_blank())

LMplotRelGrowthPost


```


### Absolute bud growth anomaly

```{r}

# Absolute bud growth anomaly
#############################

## all

LMplotAbsAutGrowth = ggplot() + 
  scale_color_manual(values = c(rep('#F21A00',3), rep('#3B9AB2',3)))+
  geom_hline(yintercept=0)+
  geom_point(data=resultsAbsAutGrowth, aes(x=term, y=estimate, color=term))+
  geom_errorbar(data=resultsAbsAutGrowth, aes(x=term, ymin=estimate-1.96*std.error, ymax=estimate+1.96*std.error, color=term), 
                width=.2, position=position_dodge(.9)) +
  
  geom_vline(xintercept=3.5, linewidth=2, alpha=0.4)+
  coord_cartesian(ylim=c(-5,5))+
  xlab("")+ylab("Absolute bud growth anomaly (mm)")+
  scale_x_discrete(label =c('Pre_day','Pre_night', 'Pre_full','Post_day', 'Post_night', 'Post_full'))+
  plotTheme1

LMplotAbsAutGrowth 


```


### Length and growth curves

```{r}

# Bud length curves
###################

budpal = c('black', 'magenta3', 'red', 'brown3', 'steelblue3', 'blue4', 'cyan2')


#move points .05 to the left and right
pd = position_dodge(2) 


# Relative bud length
RelBudLength_curve = bud.df %>%
  mutate(Treatment = factor(Treatment, 
  levels=c("Control",'Pre_day','Pre_night', 'Pre_full','Post_day', 'Post_night', 'Post_full'))) %>% 
  ggplot(aes(x=date, y=bud.rel*100, colour=Treatment, group=Treatment)) +
  
  geom_hline(yintercept = 60, colour="lightgrey")+
  geom_hline(yintercept = 70, colour="lightgrey")+
  geom_hline(yintercept = 80, colour="lightgrey")+
  geom_hline(yintercept = 90, colour="lightgrey")+
  geom_hline(yintercept = 100, colour="lightgrey")+
  
  geom_vline(xintercept = as.Date('2022-11-01'), colour="lightgrey")+
  geom_vline(xintercept = as.Date('2022-10-01'), colour="lightgrey")+
  geom_vline(xintercept = as.Date('2022-09-01'), colour="lightgrey")+
  geom_vline(xintercept = as.Date('2022-08-01'), colour="lightgrey")+
  
  stat_summary(fun.data = "mean_se", geom="line", size = 1.2, position=pd, alpha=0.7) +
  stat_summary(fun.data = "mean_se", geom="errorbar", size = 0.8, width=0, position=pd) +
  stat_summary(fun.data = "mean_se", geom="point", size = 1.2, position=pd) +
  
  scale_color_manual(values = budpal)+
  
  coord_cartesian(ylim=c(65,100),xlim=c(as.Date(c('2022-08-25','2022-11-05'))))+
  
  xlab("") +
  ylab("Relative bud length (%)") +
  
  #ggtitle("Relative bud growth")+
  
  plotTheme1 +
  theme(legend.position       = "inside",
        legend.position.inside= c(0.8, 0.5),
        legend.key            = element_rect(fill = "transparent"),
        legend.background     = element_rect(fill='white'),
        legend.box.background = element_rect(fill='transparent'),
        axis.title.x          = element_blank(),
        axis.text.x           = element_blank())

RelBudLength_curve


# Absolute bud growth
#####################


AbsBudLength_curve = bud.df %>%
  mutate(Treatment = factor(Treatment, 
    levels=c("Control",'Pre_day','Pre_night', 'Pre_full','Post_day', 'Post_night', 'Post_full'))) %>% 
  ggplot(aes(x=date, y=bud_length, colour=Treatment, group=Treatment)) +
  
  geom_hline(yintercept = 12, colour="lightgrey")+
  geom_hline(yintercept = 16, colour="lightgrey")+
  #geom_hline(yintercept = 17.5, colour="lightgrey")+
  geom_hline(yintercept = 20, colour="lightgrey")+
  geom_hline(yintercept = 24, colour="lightgrey")+
  #geom_hline(yintercept = 26, colour="lightgrey")+
  
  geom_vline(xintercept = as.Date('2022-11-01'), colour="lightgrey")+
  geom_vline(xintercept = as.Date('2022-10-01'), colour="lightgrey")+
  geom_vline(xintercept = as.Date('2022-09-01'), colour="lightgrey")+
  geom_vline(xintercept = as.Date('2022-08-01'), colour="lightgrey")+
  
  stat_summary(fun.data = "mean_se", geom="line", size = 1.2, position=pd, alpha=0.7) +
  stat_summary(fun.data = "mean_se", geom="errorbar", size = 0.8, width=0, position=pd) +
  stat_summary(fun.data = "mean_se", geom="point", size = 1.2, position=pd) +
  
  scale_color_manual(values = budpal)+
  
  coord_cartesian(ylim=c(12.5,26),xlim=c(as.Date(c('2022-08-25','2022-11-05'))))+
  
  xlab("") +
  ylab("Bud length (mm)") +
  
  #ggtitle("Absolute bud growth") +
  
  plotTheme1 

AbsBudLength_curve


# Relative bud growth rate
##########################


RelBudGrowth_curve = bud.df %>%
  ggplot(aes(x=date, y=relbud_growth*100/7, colour=Treatment, group=Treatment)) +
  
  geom_hline(yintercept = 0.25, colour="lightgrey")+
  geom_hline(yintercept = 0.5, colour="lightgrey")+
  geom_hline(yintercept = 0.75, colour="lightgrey")+
  geom_hline(yintercept = 1, colour="lightgrey")+
  
  geom_vline(xintercept = as.Date('2022-11-01'), colour="lightgrey")+
  geom_vline(xintercept = as.Date('2022-10-01'), colour="lightgrey")+
  geom_vline(xintercept = as.Date('2022-09-01'), colour="lightgrey")+
  geom_vline(xintercept = as.Date('2022-08-01'), colour="lightgrey")+
  
  stat_summary(fun.data = "mean_se", geom="line", size = 1.2, position=pd, alpha=0.7) +
  stat_summary(fun.data = "mean_se", geom="errorbar", size = 0.8, width=0, position=pd) +
  stat_summary(fun.data = "mean_se", geom="point", size = 1.2, position=pd) +
  
  scale_color_manual(values = budpal)+
  
  coord_cartesian(ylim=c(0.056,1.2),xlim=c(as.Date(c('2022-08-25','2022-11-05'))))+
  
  xlab("") +
  ylab("Relative growth per day (%)") +
  
  #ggtitle("Bud growth rate")+
  
  plotTheme1 
 

RelBudGrowth_curve


# Absolute bud growth
#####################


AbsBudGrowth_curve = bud.df %>%
  ggplot(aes(x=date, y=bud_growth/7, colour=Treatment, group=Treatment)) +
  
  geom_hline(yintercept = .1, colour="lightgrey")+
  geom_hline(yintercept = .2, colour="lightgrey")+
  geom_hline(yintercept = .3, colour="lightgrey")+
  
  geom_vline(xintercept = as.Date('2022-11-01'), colour="lightgrey")+
  geom_vline(xintercept = as.Date('2022-10-01'), colour="lightgrey")+
  geom_vline(xintercept = as.Date('2022-09-01'), colour="lightgrey")+
  geom_vline(xintercept = as.Date('2022-08-01'), colour="lightgrey")+
  
  stat_summary(fun.data = "mean_se", geom="line", size = 1.2, position=pd, alpha=0.7) +
  stat_summary(fun.data = "mean_se", geom="errorbar", size = 0.8, width=0, position=pd) +
  stat_summary(fun.data = "mean_se", geom="point", size = 1.2, position=pd) +
  
  scale_color_manual(values = budpal)+
  
  coord_cartesian(ylim=c(0.011,0.25),xlim=c(as.Date(c('2022-08-25','2022-11-05'))))+
  
  xlab("") +
  ylab("mm per day") +
  
  #ggtitle("Absolute bud growth")+
  
  plotTheme1 

AbsBudGrowth_curve

```


### Merge plots

``` {r}

# Merge Night-Full-Day pre- and post plots
#############################################

# #define plot layout
layout <- "
AB"

CombinedBudSepPlotPre_Post_NFD =  LMplot90.pre2 + LMplot90.post2 +
  plot_layout(design = layout, tag_level = 'new') + plot_annotation(tag_levels = list(c('','','')))&
  theme(plot.tag = element_text(face = 'bold'), axis.title = element_text(size = 22), plot.title = element_text(size = 22),
        axis.text = element_text(size = 18))

CombinedBudSepPlotPre_Post_NFD
#Save PDF
#pdf(paste(output.dir,"CombinedBudSepPlotPre_Post_NFD_flipped.pdf",sep="/"), width=8, height=6, useDingbats=FALSE)
#CombinedBudSepPlotPre_Post_NFD
#dev.off()

# Merge separated bud growth anomaly plots
##########################################

CombinedRelGroPlotPre_Post_NFD =  LMplotRelGrowthPre + LMplotRelGrowthPost +
  plot_layout(design = layout, tag_level = 'new') + plot_annotation(tag_levels = list(c('','','')))&
  theme(plot.tag = element_text(face = 'bold'), axis.title = element_text(size = 22), plot.title = element_text(size = 22),
        axis.text = element_text(size = 20))

CombinedRelGroPlotPre_Post_NFD
#Save PDF
#pdf(paste(output.dir,"CombinedRelGroPlotPre_Post_NFD.pdf",sep="/"), width=8, height=6, useDingbats=FALSE)
#CombinedRelGroPlotPre_Post_NFD
#dev.off()

```


### Individual bud length curves

``` {r}

budlength_individuals = bud.df %>%
  ggplot(aes(x=date, y=bud_length, 
             colour=Treatment, fill=Treatment)) +
  
  geom_vline(xintercept = as.Date('2022-11-01'), colour="lightgrey")+
  geom_vline(xintercept = as.Date('2022-10-01'), colour="lightgrey")+
  geom_vline(xintercept = as.Date('2022-09-01'), colour="lightgrey")+
  geom_vline(xintercept = as.Date('2022-08-01'), colour="lightgrey")+
  geom_vline(xintercept = as.Date('2022-07-01'), colour="lightgrey")+
  
  geom_hline(yintercept = 10, colour="lightgrey")+
  geom_hline(yintercept = 20, colour="lightgrey")+
  geom_hline(yintercept = 30, colour="lightgrey")+
  
  geom_line(aes(x=date, y=bud_length), color="black")+
  geom_point()+
  geom_area(aes(x=date, y=bud_length), alpha=0.5) +
  
  geom_vline(data=bud.data.90, 
             aes(xintercept = EOS), linetype = "dashed")+
  
  xlab("Date") +
  ylab("Bud length (mm)") +
  
  facet_grid(ID~bud_type)+
  plotTheme1

budlength_individuals

#Save PDF
#pdf(paste(output.dir,"IndividualPlot_Budlength.pdf",sep="/"), width=6, height=120, useDingbats=FALSE)
#budlength_individuals
#dev.off()

```


## Reproducibility

```{r}
sessionInfo()
```