---
title: "Development underpins the summer solstice reversal of the effects of climate warming on the autumn phenology of European beech"
subtitle: "Analysis of leaf senescence data for Experiment 2"
author: "Dominic Rebindaine"
date: "2025-08-05"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction

This document displays the analysis of leaf chlorophyll data of experiment 2 in "Development underpins the summer solstice reversal of the effects of climate warming on the autumn phenology of European beech" Rebindaine et al. (2025). The data is loaded and restructured, leaf senescence dates (end-of-season, EOS) are derived and analysed in context of the various treatments, then figures are plotted.

### Load required packages

```{r, message=FALSE}
library(tidyverse)
library(cropgrowdays)
library(lmerTest)
require(broom.mixed)
require(patchwork)
require(data.table)
require(lme4)
require(eoffice)
require(ggpmisc)
require(parameters)
```

### Configure input/output

```{r in/out}

input.path = '/Users/dominic/Library/Mobile Documents/com~apple~CloudDocs/Documents/PhD/development_2022_2023/analysis/experiment_2/Data'

output.path = '/Users/dominic/Library/Mobile Documents/com~apple~CloudDocs/Documents/PhD/development_2022_2023/analysis/experiment_2/figures/leaves'

```

### Load and restructure data

```{r load data}
# import data
chlor.df = read.csv(paste(input.path, '2022_phen_exp_spad.csv', sep = '/'))

chlor.df = chlor.df %>% 
  mutate(
    #date formatting
    measurement.date = as.Date(measurement.date, format = '%d.%m.%Y'),
    #Convert dates to DOY
    measurement.doy = day_of_year(measurement.date),
        # turn treatment groups into meaningful terms and merge controls
        Treatment = ifelse(group %in% c("A","C", "D","H"),"Control",
                            ifelse(group %in% c("E"),"Pre_day", 
                                ifelse(group %in% c("F"),"Pre_night",
                                   ifelse(group %in% c("G"),"Pre_full", 
                                          ifelse(group %in% c("I"),"Post_day",
                                                 ifelse(group %in% c("J"),"Post_night",
                                   ifelse(group %in% c("K"), "Post_full", "Other"))))))) 
) %>% 
  
    # remove B treatment, and NAs
  filter(!Treatment %in% c('Other'),
         !is.na(SPAD)) %>% 
    # group to individual
  group_by(id) %>% 
   # only keep trees with more than 8 measurements
  filter(n() >= 8) %>% 
  ungroup()
   
     # factorise Treatment and relevel
  chlor.df$Treatment = as.factor(chlor.df$Treatment)
  chlor.df$Treatment = fct_relevel(chlor.df$Treatment, 'Control', 'Pre_night', 
                                   'Pre_full', 'Pre_day', 'Post_night', 'Post_full', 'Post_day')

str(chlor.df)


# Convert SPAD to chlorophyll content
SPADtoChl = function(SPAD){
  Chl = ifelse(SPAD==0, 0, -0.0029*(SPAD^2) + 1.175*SPAD + 3.8506)
  return(Chl)
}
#SPAD: unitless
#Chl: leaf chlorophyll in microgram / gram fresh weight
#Intercept forced to zero

#From Percival et al. 2008: The Potential of a Chlorophyll Content SPAD Meter to Quantify Nutrient Stress in Foliar Tissue of Sycamore (Acer pseudoplatanus), English Oak (Quercus robur), and European Beech (Fagus sylvatica)

chlor.df = chlor.df %>% 
  mutate(chlorophyll = SPADtoChl(SPAD))

# get mean and sd of groups over time
sum_chlor = chlor.df %>% 
  group_by(group, measurement.doy, measurement.date) %>% 
  summarise(across(chlorophyll, list(
        mean = \(x) mean(x, na.rm = TRUE), 
        sd = \(x) sd(x, na.rm = TRUE))),
    .groups = 'keep')

```

### Outlier detection
Find measurements that are more than 1.5 times the IQR outside of the lowest/highest quartiles. 
Quartiles and IQR are computed within treatment and emasurement date.

```{r}

# Function to flag outliers
flag_iqr_outliers = function(x) {
  q1 = quantile(x, 0.25, na.rm = TRUE)
  q3 = quantile(x, 0.75, na.rm = TRUE)
  iqr = q3 - q1
  lower = q1 - 1.5 * iqr
  upper = q3 + 1.5 * iqr
  return(x < lower | x > upper)
}

# Apply the function per treatment and date
chlor_outliers_flagged = chlor.df %>%
  group_by(Treatment, measurement.date) %>%
  mutate(outlier = flag_iqr_outliers(chlorophyll)) %>%
  ungroup()

# visualise outliers

ggplot(chlor_outliers_flagged, aes(x = measurement.date, y = chlorophyll)) +
  geom_point(aes(color = outlier)) +
  geom_text(
    data = subset(chlor_outliers_flagged, outlier == TRUE),
    aes(label = id),
    vjust = -0.5,        # adjust vertical position
    size = 3,            # text size
    color = "black",     # override point color for visibility
    check_overlap = TRUE # avoid overlapping labels
  ) +
  facet_wrap(~ Treatment) +
  labs(
    title = "Outliers by treatment and measurement date (IQR Method)") +
  theme_minimal()

##### Remove trees with > 1 outlier measurement and then delete outlier points

# Count outliers per tree
tree_outlier_counts = chlor_outliers_flagged %>%
  group_by(id) %>%
  summarise(n_outliers = sum(outlier, na.rm = TRUE)) %>%
  ungroup()

# Identify trees with more than 1 outlier
trees_to_remove = tree_outlier_counts %>%
  filter(n_outliers > 1) %>%
  pull(id)

# Remove those trees from the dataset
chlor.df = chlor_outliers_flagged %>%
  filter(!(id %in% trees_to_remove) & !outlier) %>% 
    select(!outlier)

message("Removed trees with >1 outlier: ", paste(trees_to_remove, collapse = ", "))

```

### Sample sizes

```{r sample sizes}

for (date in unique(chlor.df$measurement.date)) {
  
  print(date)
  # Observed trees per treatment
print(table(chlor.df[chlor.df$measurement.date== date,]$Treatment))
  
}

# Number of individual trees
length(unique(chlor.df$id))
```

### Relative chlorophyll levels

```{r rel chlor}

chlor.df = chlor.df %>%
  group_by(id)%>%
  mutate(
    #get relative chlorophyll content per individual
    chlorophyll.rel = chlorophyll / max(chlorophyll),
    #save max chlorophyll content for each individual
    chlorophyll.max = chlorophyll.rel == max(chlorophyll.rel),
    #get date of maximum chlorophyll content per individual
    date.max.chl = measurement.date[which.max(chlorophyll)]) %>%
  ungroup()

```


### EOS (leaf senescence timing) derivation

```{r EOS}

## Interpolate 10%, 25%, 50% and 85% senescence dates from seasonal chlorophyll measurements

# Interpolate EOS10 (date chlorophyll last dropped below 90%)
senescence.data.10 = chlor.df %>%
  group_by(id, Treatment) %>%
  # Find the last row with >90% chlorophyll in each group, keep only this row and next
  filter(row_number() <= max(which(chlorophyll.rel > 0.9))+1,
         row_number() >= max(which(chlorophyll.rel > 0.9))) %>%
  #Linear interpolation
  dplyr::summarize(EOS     = as.Date(approx(chlorophyll.rel, measurement.date, .9,  ties=min)$y, origin="1970-01-01")) %>%
  mutate(EOS.DOY    = yday(EOS),
         phenophase = "EOS10") %>%
  ungroup()

# Interpolate EOS25 (date chlorophyll last dropped below 75%)
senescence.data.25 = chlor.df %>%
  group_by(id, Treatment) %>%
  # Find the last row with >75% chlorophyll in each group, keep only this row and next
  filter(row_number() <= max(which(chlorophyll.rel > 0.75))+1,
         row_number() >= max(which(chlorophyll.rel > 0.75))) %>%
  #Linear interpolation
  dplyr::summarize(EOS     = as.Date(approx(chlorophyll.rel, measurement.date, .75,  ties=min)$y, origin="1970-01-01")) %>%
  mutate(EOS.DOY    = yday(EOS),
         phenophase = "EOS25") %>%
  ungroup()

# Interpolate EOS50 (date chlorophyll last dropped below 50%)
senescence.data.50 = chlor.df %>%
  group_by(id, Treatment) %>%
  # Find the last row with >50% chlorophyll in each group, keep only this row and next
  filter(row_number() <= max(which(chlorophyll.rel > 0.5))+1,
         row_number() >= max(which(chlorophyll.rel > 0.5))) %>%
  #Linear interpolation
  dplyr::summarize(EOS     = as.Date(approx(chlorophyll.rel, measurement.date, .5,  ties=min)$y, origin="1970-01-01")) %>%
  mutate(EOS.DOY    = yday(EOS),
         phenophase = "EOS50") %>%
  ungroup()

# Interpolate EOS85 (date chlorophyll last dropped below 85%)
senescence.data.85 = chlor.df %>%
  group_by(id, Treatment) %>%
  # Find the last row with >85% chlorophyll in each group, keep only this row and next
  filter(row_number() <= max(which(chlorophyll.rel > 0.15))+1,
         row_number() >= max(which(chlorophyll.rel > 0.15))) %>%
  #Linear interpolation
  dplyr::summarize(EOS     = as.Date(approx(chlorophyll.rel, measurement.date, .15,  ties=min)$y, origin="1970-01-01")) %>%
  mutate(EOS.DOY    = yday(EOS),
         phenophase = "EOS85") %>%
  ungroup()

# Rbind
senescence.data = rbind(senescence.data.10,senescence.data.25,senescence.data.50, senescence.data.85)

# save data
#write_csv(senescence.data, file = paste(input.path, "senescence_data_experiment2.csv", sep = "/"))

# Mean EOS DOY per treatment
##############################

eos.doy.mean = 
  senescence.data %>% 
  group_by(Treatment, phenophase) %>% 
  summarise(meanEOS.DOY = round(mean(EOS.DOY),1),
            ci.lower.EOS = round(meanEOS.DOY - 1.96*standard_error(EOS.DOY),1),
            ci.upper.EOS = round(meanEOS.DOY + 1.96*standard_error(EOS.DOY),1))



```


### ANOVAs by treatment

```{r}

# vector of phenophases
phenophases = unique(senescence.data$phenophase)

# for loop over all phenophases
for (phenophase in phenophases) {
  
  print(phenophase)
  
  model.anova.phenophase = senescence.data %>% 
    filter(phenophase == phenophase) %>% 
      aov(EOS.DOY ~ Treatment, data=.)
  
  result.anova.phenophase = summary(model.anova.phenophase)
  
  print(result.anova.phenophase)
  
}

```


## EOS Analysis

### EOS50

```{r leaf senescence analysis}

### Linear modelling of leaf senescence timing 

## Treatment as only  predictor variable
resultsLM.50 = senescence.data.50 %>% 
  do({model = lm(EOS.DOY ~ Treatment, data=.)  # create the model
  data.frame(tidy(model))}) %>%
  filter(!term %in% c("(Intercept)")) 

resultsLM.50 = resultsLM.50 %>% 
  mutate(term = factor(term, levels = c('TreatmentPre_night','TreatmentPre_full','TreatmentPre_day', 
  'TreatmentPost_night', 'TreatmentPost_full', 'TreatmentPost_day')),
  SEx2 = std.error*2)
       
#Show results
as.data.frame(resultsLM.50) %>%
  dplyr::select(-c(statistic))%>%
  mutate_if(is.numeric, round, digits=2)

```

## Publication Figures

#### Plot theme

```{r plot theme}

  plotTheme1 = theme(
    legend.position   = "none",
    legend.background = element_rect(fill=NA, linewidth=0.5, linetype="solid"),
    legend.text       = element_text(color="black"),
    panel.background  = element_blank(),
    axis.text         = element_text(colour = "black"),
    panel.border      = element_rect(colour = "black", fill=NA),
    axis.line         = element_line(color = "black"),
    strip.background  = element_rect(fill=NA),
    strip.text        = element_text(colour = 'black'),
    plot.title        = element_text(face="bold",hjust = 0.5))
  
  chlorpal = c('#a6cee3','#1f78b4','#b2df8a','#33a02c','#fb9a99','#e31a1c','#fdbf6f','#ff7f00','#cab2d6','#6a3d9a')
  

```


### Leaf senescence figures

#### EOS50

```{r leaf sen pub figs}

# Plot mirrored across y = 0
################################################


## Pre-solstice cooling

LMplot50.pre = ggplot() + 
  scale_color_manual(values = c('#F21A00', '#F21A00','#F21A00'))+
  geom_hline(yintercept=0)+
  geom_point(data= filter(resultsLM.50, term %in% c('TreatmentPre_night', 'TreatmentPre_full', 'TreatmentPre_day')), 
             aes(x=term, y=estimate*-1, color=term), size = 3.5)+
  geom_errorbar(data=filter(resultsLM.50, term %in% c('TreatmentPre_night', 'TreatmentPre_full', 'TreatmentPre_day')), 
                aes(x=term, ymin=-1*estimate-1.96*std.error, ymax=-1*estimate+1.96*std.error, color=term), linewidth = 1, 
                width=.2, position=position_dodge(.9)) +
  coord_cartesian(ylim=c(-10,10))+
  scale_x_discrete(labels = c('Night','Full-day', 'Day'))+
  xlab("")+ylab("EOS50 advancement (days)")+
  plotTheme1 +
  ggtitle('Pre-solstice cooling')

LMplot50.pre

## Post-solstice cooling

LMplot50.post = ggplot() + 
  scale_color_manual(values = c('#3B9AB2', '#3B9AB2','#3B9AB2'))+
  geom_hline(yintercept=0)+
  geom_point(data= filter(resultsLM.50, term %in% c('TreatmentPost_night', 'TreatmentPost_full', 'TreatmentPost_day')), 
             aes(x=term, y=estimate*-1, color=term), size = 3.5)+
  geom_errorbar(data=filter(resultsLM.50, term %in% c('TreatmentPost_night', 'TreatmentPost_full', 'TreatmentPost_day')), 
                aes(x=term, ymin=-1*estimate-1.96*std.error, ymax=-1*estimate+1.96*std.error, color=term), linewidth = 1,
                width=.2, position=position_dodge(.9)) +
  coord_cartesian(ylim=c(-10,10))+
  xlab("")+ylab("EOS50 advancement (days)")+
  scale_x_discrete(labels = c('Night','Full-day', 'Day'))+
  plotTheme1 +
  ggtitle('Post-solstice cooling') +
  theme(axis.title.y=element_blank(),
        axis.text.y=element_blank())

LMplot50.post

```

#### Merge EOS50 plots
```{r merge plots}

# Merge Night-Full-Day pre- and post plots
#############################################

# #define plot layout
layout <- "
AB"

CombinedLeafSepPlotPre_Post_NFD =  LMplot50.pre + LMplot50.post +
  plot_layout(design = layout, tag_level = 'new') + plot_annotation(tag_levels = list(c('','','')))&
  theme(plot.tag = element_text(face = 'bold'), axis.title = element_text(size = 22), plot.title = element_text(size = 22),
        axis.text = element_text(size = 18))

#Save PDF
#pdf(paste(output.path,"CombinedLeafSepPlotPre_Post_NFD_flipped.pdf",sep="/"), width=8, height=6, useDingbats=FALSE)
#CombinedLeafSepPlotPre_Post_NFD
#dev.off()



```

### Relative chlor figures

#### By group

```{r}
# chlor curves
###################

#move points .05 to the left and right
pd = position_dodge(2) 


# Relative chlor 
Relchlor_curve = chlor.df %>%
  ggplot(aes(x=measurement.date, y=chlorophyll.rel*100, colour=Treatment, group=Treatment)) +
  
  geom_hline(yintercept = 15, colour="lightgrey")+
  geom_hline(yintercept = 50, colour="lightgrey")+
  geom_hline(yintercept = 75, colour="lightgrey")+
  geom_hline(yintercept = 90, colour="lightgrey")+
  geom_hline(yintercept = 100, colour="lightgrey")+
  
  geom_vline(xintercept = as.Date('2022-11-01'), colour="lightgrey")+
  geom_vline(xintercept = as.Date('2022-10-01'), colour="lightgrey")+
  geom_vline(xintercept = as.Date('2022-09-01'), colour="lightgrey")+
  geom_vline(xintercept = as.Date('2022-08-01'), colour="lightgrey")+
  
  stat_summary(fun.data = "mean_se", geom="line", size = 1.2, position=pd, alpha=0.7) +
  stat_summary(fun.data = "mean_se", geom="errorbar", size = 0.8, width=0, position=pd) +
  stat_summary(fun.data = "mean_se", geom="point", size = 1.2, position=pd) +
  
  scale_color_manual(values = chlorpal)+
  
  coord_cartesian(ylim=c(0,100),xlim=c(as.Date(c('2022-06-20','2022-12-16'))))+
  
  xlab("") +
  ylab("Relative chlorophyll content (%)") +
  
  #ggtitle("Relative chlor ")+
  
  plotTheme1

Relchlor_curve

```

#### By individual

```{r}

# plot relative chlor length for all individs
individ.relchlor.curves = chlor.df %>% ggplot(
  aes(x=measurement.date,y=chlorophyll.rel*100,colour = group, fill = group)) +
  
  geom_vline(xintercept = as.Date('2022-11-01'), colour="lightgrey")+
  geom_vline(xintercept = as.Date('2022-10-01'), colour="lightgrey")+
  geom_vline(xintercept = as.Date('2022-09-01'), colour="lightgrey")+
  geom_vline(xintercept = as.Date('2022-08-01'), colour="lightgrey")+
  geom_vline(xintercept = as.Date('2022-07-01'), colour="lightgrey")+
  
  geom_hline(yintercept = 15, colour="lightgrey")+
  geom_hline(yintercept = 50, colour="lightgrey")+
  geom_hline(yintercept = 75, colour="lightgrey")+
  geom_hline(yintercept = 90, colour="lightgrey")+
  geom_hline(yintercept = 100, colour="lightgrey")+

  
  geom_line(aes(x=measurement.date, y=chlorophyll.rel*100), color="black")+
  geom_point()+
  
  geom_vline(data=senescence.data.50, 
             aes(xintercept = EOS), linetype = "dashed", colour = 'red')+
   geom_vline(data=senescence.data.10, 
             aes(xintercept = EOS), linetype = "dashed", colour = 'green')+
   geom_vline(data=senescence.data.25, 
             aes(xintercept = EOS), linetype = "dashed", colour = 'blue')+
   geom_vline(data=senescence.data.85, 
             aes(xintercept = EOS), linetype = "dashed", colour = 'brown')+
  
  coord_cartesian(ylim=c(0,110),xlim=c(as.Date(c('2022-06-20','2022-12-16'))))+
  
  xlab("Date") +
  ylab("Relative chlor (%)") +
    facet_wrap(~id, dir='v', nrow = 12) +
    plotTheme1 +
  theme(legend.position = "none")
  
individ.relchlor.curves
  #Save PDF
#ggsave(paste(output.path, "individv_chlor_curves_no_removals.pdf", sep = "/"), height = 40, width = 40)
  

```

## Reproducibility

```{r}

sessionInfo()

```
